---
title: "Main"
format: html
editor: visual
---

## Packages

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(foreign)
library(janitor)
library(httr)
library(glue)
library(zip)
library(GGally)

# Suppress scientific notation in all output
options(scipen = 999)
```

## Loading And Preparing the Data

```{r}
#| message: false
# Load pre-cleaned TidyTuesday versions as a fallback reference
exped_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/exped_tidy.csv')
peaks_tidy <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/peaks_tidy.csv')
```

Provided cleaning script:

```{r}
#| eval: false
#| message: false
#| warning: false
################################################################################
### Cleaning data from the Himalayan Database ##################################
################################################################################

###_____________________________________________________________________________
### Utilize httr and foreign to gain access to the file and import
###_____________________________________________________________________________

# Step 1: Set up the URL and file paths
url <- "https://www.himalayandatabase.com/downloads/Himalayan%20Database.zip"
zip_file <- tempfile(fileext = ".zip")
extracted_dir <- tempfile()

# Step 2: Download and extract the ZIP file
httr::GET(url, httr::write_disk(zip_file, overwrite = TRUE))
zip::unzip(zip_file, exdir = extracted_dir)

# Step 3: Read the .DBF files
dbf_file_peaks <- file.path(extracted_dir, "HIMDATA", "peaks.DBF")
dbf_file_exped <- file.path(extracted_dir, "HIMDATA", "exped.DBF")
peaks_data <- foreign::read.dbf(dbf_file_peaks, as.is = FALSE)
exped_data <- foreign::read.dbf(dbf_file_exped, as.is = FALSE)

# Step 4: Convert to tidy tibble
peaks_temp <- tibble::as_tibble(peaks_data)
exped_temp <- tibble::as_tibble(exped_data)

# Clean up temporary files (optional)
unlink(c(zip_file, extracted_dir), recursive = TRUE)

###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the peaks dataset
###_____________________________________________________________________________

peaks_tidy <- peaks_temp |> 
  dplyr::mutate(HIMAL_FACTOR = case_when(
    HIMAL == 0  ~ "Unclassified",
    HIMAL == 1  ~ "Annapurna",
    HIMAL == 2  ~ "Api/Byas Risi/Guras",
    HIMAL == 3  ~ "Damodar",
    HIMAL == 4  ~ "Dhaulagiri",
    HIMAL == 5  ~ "Ganesh/Shringi",
    HIMAL == 6  ~ "Janak/Ohmi Kangri",
    HIMAL == 7  ~ "Jongsang",
    HIMAL == 8  ~ "Jugal",
    HIMAL == 9  ~ "Kangchenjunga/Simhalila",
    HIMAL == 10 ~ "Kanjiroba",
    HIMAL == 11 ~ "Kanti/Palchung",
    HIMAL == 12 ~ "Khumbu",
    HIMAL == 13 ~ "Langtang",
    HIMAL == 14 ~ "Makalu",
    HIMAL == 15 ~ "Manaslu/Mansiri",
    HIMAL == 16 ~ "Mukut/Mustang",
    HIMAL == 17 ~ "Nalakankar/Chandi/Changla",
    HIMAL == 18 ~ "Peri",
    HIMAL == 19 ~ "Rolwaling",
    HIMAL == 20 ~ "Saipal",
    TRUE ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = HIMAL
  ) |> 
  mutate(REGION_FACTOR = case_when(
    REGION == 0 ~ "Unclassified",
    REGION == 1 ~ "Kangchenjunga-Janak",
    REGION == 2 ~ "Khumbu-Rolwaling-Makalu",
    REGION == 3 ~ "Langtang-Jugal",
    REGION == 4 ~ "Manaslu-Ganesh",
    REGION == 5 ~ "Annapurna-Damodar-Peri",
    REGION == 6 ~ "Dhaulagiri-Mukut",
    REGION == 7 ~ "Kanjiroba-Far West",
    TRUE        ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = REGION
  ) |> 
  mutate(PHOST_FACTOR = case_when(
    PHOST == 0 ~ "Unclassified",
    PHOST == 1 ~ "Nepal only",
    PHOST == 2 ~ "China only",
    PHOST == 3 ~ "India only",
    PHOST == 4 ~ "Nepal & China",
    PHOST == 5 ~ "Nepal & India",
    PHOST == 6 ~ "Nepal, China & India",
    TRUE       ~ "Unknown" # Default case if a value doesn't match
  ),
  .after = PHOST
  ) |> 
  mutate(PSTATUS_FACTOR = case_when(
    PSTATUS == 0 ~ "Unknown",
    PSTATUS == 1 ~ "Unclimbed",
    PSTATUS == 2 ~ "Climbed",
    TRUE         ~ "Invalid" # Default case if a value doesn't match
  ),
  .after = PSTATUS
  ) |> 
  mutate(across(c(HIMAL_FACTOR, PHOST_FACTOR, PSTATUS_FACTOR), ~ factor(.)))

###_____________________________________________________________________________
### Add columns that recode some variables that are integers into character
### strings that describe their subject in the expeditions dataset
###_____________________________________________________________________________

exped_tidy <- exped_temp |> 
  dplyr::mutate(SEASON_FACTOR = case_when(SEASON == 0 ~ "Unknown",
                                          SEASON == 1 ~ "Spring",
                                          SEASON == 2 ~ "Summer",
                                          SEASON == 3 ~ "Autumn",
                                          SEASON == 4 ~ "Winter",
                                          TRUE ~ NA_character_
                                          ),
                .after = SEASON
                ) |> 
  dplyr::mutate(HOST_FACTOR = case_when(HOST == 0 ~ "Unknown",
                                        HOST == 1 ~ "Nepal",
                                        HOST == 2 ~ "China",
                                        HOST == 3 ~ "India",
                                        TRUE ~ NA_character_
                                        ),
                .after = HOST
                ) |> 
  dplyr::mutate(TERMREASON_FACTOR = case_when(TERMREASON == 0 ~ "Unknown",
                                       TERMREASON == 1 ~ "Success (main peak)",
                                       TERMREASON == 2 ~ "Success (subpeak, foresummit)",
                                       TERMREASON == 3 ~ "Success (claimed)",
                                       TERMREASON == 4 ~ "Bad weather (storms, high winds)",
                                       TERMREASON == 5 ~ "Bad conditions (deep snow, avalanching, falling ice, or rock)",
                                       TERMREASON == 6 ~ "Accident (death or serious injury)",
                                       TERMREASON == 7 ~ "Illness, AMS, exhaustion, or frostbite",
                                       TERMREASON == 8 ~ "Lack (or loss) of supplies, support or equipment",
                                       TERMREASON == 9 ~ "Lack of time",
                                       TERMREASON == 10 ~ "Route technically too difficult, lack of experience, strength, or motivation",
                                       TERMREASON == 11 ~ "Did not reach base camp",
                                       TERMREASON == 12 ~ "Did not attempt climb",
                                       TERMREASON == 13 ~ "Attempt rumored",
                                       TERMREASON == 14 ~ "Other"
                                       ),
                .after = TERMREASON
                ) |> 
  filter(grepl(pattern = "202[01234]", x = YEAR))

################################################################################
### End  #######################################################################
################################################################################
```

## Exploration

```{r}
# Top 10 peaks by number of successful summits (TERMREASON == 1: main peak only)
success_count = exped_tidy |>
  filter(TERMREASON == 1) |>
  count(PEAKID) |>
  arrange(desc(n)) |>
  slice_head(n = 10)

ggplot(success_count, aes(x = reorder(PEAKID, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Peaks by Successful Summits (2020-2024)",
       x = "Peak ID", y = "# of Successful Summits")
```

```{r}
# Top 10 peak-year combinations by total attempts
# Helps show which peaks are most active and in which years
year_count = exped_tidy |>
  count(PEAKID, YEAR) |>
  arrange(desc(n)) |>
  slice_head(n = 10)

ggplot(year_count, aes(x = interaction(PEAKID, YEAR), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Top 10 Peak-Year Combinations by Attempts", 
       x = "Peak ID . Year", y = "Number of Attempts") +
  coord_flip()
```

```{r}
# Peaks where supplemental oxygen was most commonly used
# Oxygen use is concentrated on the highest peaks (e.g. Everest, Lhotse)
oxy_count = exped_tidy |>
  filter(O2USED == TRUE) |>
  count(PEAKID) |>
  arrange(desc(n)) |>
  slice_head(n = 10)

ggplot(oxy_count, aes(x = reorder(PEAKID, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Peaks by Oxygen Usage (2020-2024)",
       x = "Peak ID", y = "# of Attempts with O2 Usage")
```

```{r}
exped_tidy |>
  mutate(SUCCESS = as.integer(TERMREASON %in% 1:3)) |>
  group_by(SEASON_FACTOR) |>
  summarise(success_rate = mean(SUCCESS), n = n()) |>
  filter(!is.na(SEASON_FACTOR)) |>
  mutate(color_group = ifelse(success_rate >= median(success_rate), "high", "low")) |>
  ggplot(aes(x = reorder(SEASON_FACTOR, -success_rate), 
             y = success_rate, 
             fill = color_group)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("high" = "steelblue", "low" = "tomato"), 
                    guide = "none") +
  geom_text(aes(label = paste0(round(success_rate * 100, 1), "%")),
            vjust = -0.5, size = 5) +
  labs(title = "Success Rate by Season",
       x = "Season", y = "Success Rate") +
  ylim(0, 1)

```

## Height vs Success Rate (Exploratory)

```{r}
exped_tidy |>
  left_join(peaks_tidy |> select(PEAKID, HEIGHTM), by = "PEAKID") |>
  mutate(SUCCESS = as.integer(TERMREASON %in% 1:3)) |>
  filter(!is.na(HEIGHTM)) |>
  mutate(height_bin = cut(HEIGHTM, breaks = seq(5000, 9000, by = 500),
                          include.lowest = TRUE)) |>
  group_by(height_bin) |>
  summarise(success_rate = mean(SUCCESS), n = n()) |>
  filter(!is.na(height_bin)) |>
  ggplot(aes(x = height_bin, y = success_rate)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0("n=", n)), vjust = -0.4, size = 3) +
  labs(title = "Success Rate by Peak Height Band",
       x = "Height (m)", y = "Success Rate") +
  ylim(0, 1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
exped_tidy |>
  left_join(peaks_tidy |> select(PEAKID, HEIGHTM), by = "PEAKID") |>
  mutate(SUCCESS = as.integer(TERMREASON %in% 1:3)) |>
  filter(!is.na(HEIGHTM)) |>
  ggplot(aes(x = HEIGHTM, y = SUCCESS)) +
  geom_smooth(method = "loess", color = "steelblue", se = TRUE) +
  labs(title = "LOESS Smooth: Success vs Peak Height",
       x = "Peak Height (m)", y = "Predicted Success Rate")
```

## Model Fitting

```{r}
# Build model dataset:
# - SUCCESS: 1 if any form of summit success (TERMREASON 1, 2, or 3), 0 otherwise
# - Join in peak height from peaks_tidy
exped_model = exped_tidy |>
  mutate(SUCCESS = as.integer(TERMREASON %in% 1:3)) |>
  subset(select = c(PEAKID, SUCCESS, TERMREASON, O2USED, SEASON_FACTOR, CAMPS, TOTMEMBERS))

temp = subset.data.frame(peaks_tidy, select = c(PEAKID, HEIGHTM))
exped_model = left_join(exped_model, temp, by = "PEAKID") |>
  mutate(HEIGHTM_KM = HEIGHTM / 1000,
         HEIGHTM_KM_C = HEIGHTM_KM - mean(HEIGHTM_KM, na.rm = TRUE))

# Logistic regression: predicting binary summit success
# O2USED * HEIGHTM_KM includes both main effects AND interaction term
# Interaction tests whether oxygen benefit grows with altitude (central hypothesis)
model = glm(SUCCESS ~ O2USED * HEIGHTM_KM_C + I(HEIGHTM_KM_C^2) + SEASON_FACTOR + CAMPS + TOTMEMBERS,
            data = exped_model, family = "binomial")
summary(model)
```

```{r}
# Summary stats by oxygen use — useful context for the presentation
# Shows success rate and average peak height for O2 vs non-O2 expeditions
exped_model |>
  group_by(O2USED) |>
  summarise(
    n = n(),
    success_rate = round(mean(SUCCESS), 3),
    avg_height = round(mean(HEIGHTM_KM, na.rm = TRUE), 3)
  )
```

```{r}
exped_model |>
  group_by(O2USED) |>
  summarise(success_rate = mean(SUCCESS), n = n()) |>
  mutate(O2_label = ifelse(O2USED == TRUE, "O2 Used", "No O2")) |>
  ggplot(aes(x = O2_label, y = success_rate)) +
  geom_col(width = 0.4, fill = c("steelblue", "tomato")) +
  geom_text(aes(label = paste0(round(success_rate * 100, 1), "%")),
            vjust = -0.5, size = 5) +
  labs(title = "Success Rate by Oxygen Use",
       x = "Oxygen Use", y = "Success Rate") +
  ylim(0, 1)

```

## Interaction Term Justification

```{r}
# Additive model: polynomial height but no O2USED:HEIGHTM_KM interaction
model_additive <- glm(SUCCESS ~ O2USED + HEIGHTM_KM_C + I(HEIGHTM_KM_C^2) + SEASON_FACTOR + CAMPS + TOTMEMBERS,
                      data = exped_model, family = "binomial")

# Likelihood ratio test: does adding the interaction significantly improve fit?
anova(model_additive, model, test = "Chisq")
```

```{r}
# AIC comparison: lower AIC = better fit penalized for model complexity
AIC(model_additive, model)
```

## Diagnostics

```{r}
# Type II ANOVA with Chi-square test — shows which predictors significantly
# reduce deviance. More appropriate than default anova() for logistic regression.
anova(model, test = "Chisq")
```

Residual diagnostics help check whether the fitted probabilities leave any
clear structure unexplained. For a logistic regression, the residuals do not need
to look normally distributed, but they should be centered around zero without a
strong trend across fitted probabilities.

```{r}
resid_diag = exped_model |>
  mutate(
    fitted_prob = fitted(model),
    deviance_resid = residuals(model, type = "deviance"),
    pearson_resid = residuals(model, type = "pearson")
  )

ggplot(resid_diag, aes(x = fitted_prob, y = deviance_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(
    title = "Deviance Residuals vs Fitted Probabilities",
    x = "Fitted probability of summit success",
    y = "Deviance residual"
  )
```

```{r}
resid_diag |>
  mutate(prob_decile = ntile(fitted_prob, 10)) |>
  group_by(prob_decile) |>
  summarise(
    mean_fitted = mean(fitted_prob),
    mean_pearson_resid = mean(pearson_resid),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = mean_fitted, y = mean_pearson_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_line(color = "steelblue") +
  geom_point(aes(size = n), color = "steelblue") +
  guides(size = "none") +
  labs(
    title = "Binned Pearson Residuals",
    x = "Average fitted probability by decile",
    y = "Average Pearson residual"
  )
```

The residual plots are mostly centered around zero, which suggests that the
model captures the main success-probability pattern. There is some remaining
structure in the middle fitted-probability range, so unmeasured factors such as
route difficulty, short-term weather, expedition support, and climber experience
may still explain part of the outcome.

```{r}
#| message: false
# Pairwise relationship plot for numeric predictors and O2USED
# Helps check for collinearity and visualize predictor distributions
temp = subset.data.frame(exped_model, select = c(O2USED, CAMPS, TOTMEMBERS, HEIGHTM_KM))
ggpairs(temp)
```

```{r}
exped_model |>
  mutate(SUCCESS = factor(SUCCESS, labels = c("Failure", "Success"))) |>
  ggparcoord(
    columns = which(colnames(exped_model) %in% c("CAMPS", "TOTMEMBERS", "HEIGHTM_KM")),
    groupColumn = which(colnames(
      mutate(exped_model, SUCCESS = factor(SUCCESS))) == "SUCCESS"),
    alphaLines = 0.3,
    scale = "uniminmax"  # scales each variable to 0-1 so axes are comparable
  ) +
  scale_color_manual(values = c("Failure" = "tomato", "Success" = "steelblue")) +
  labs(title = "Parallel Coordinates: Success vs Key Predictors",
       color = "Outcome", y = "Scaled Value")

```

```{r}
# Odds Ratio table with 95% confidence intervals
# OR > 1: increases odds of success; OR < 1: decreases odds
# NOTE: AUTUMN is the baseline season — all season ORs are relative to Autumn
# An OR confidence interval that does not cross 1 is statistically significant
round(exp(cbind(OR = coef(model), confint(model))), 3)
```

## Poly Test

```{r}
# Linear model (no polynomial) as baseline for comparison
model_linear = glm(SUCCESS ~ O2USED * HEIGHTM_KM_C + SEASON_FACTOR + CAMPS + TOTMEMBERS,
                   family = "binomial", data = exped_model)

# LRT: significant p-value confirms quadratic term improves fit
anova(model_linear, model, test = "Chisq")

# AIC: polynomial model should score lower
AIC(model_linear, model)
```

## Interaction Visualization

```{r}
# Create a prediction grid across the full range of peak heights
# Hold all other variables at their mean/reference level to isolate the interaction
# This shows how the O2 benefit changes as altitude increases
pred_grid = expand.grid(
  HEIGHTM_KM_C = seq(5.4, 8.85, by = 0.05) - mean(exped_model$HEIGHTM_KM, na.rm = TRUE),
  O2USED = c(TRUE, FALSE),
  SEASON_FACTOR = "Autumn",        # reference level
  CAMPS = mean(exped_model$CAMPS, na.rm = TRUE),
  TOTMEMBERS = mean(exped_model$TOTMEMBERS, na.rm = TRUE)
)
pred_grid$HEIGHTM_KM = pred_grid$HEIGHTM_KM_C + mean(exped_model$HEIGHTM_KM, na.rm = TRUE)
pred_grid$prob = predict(model, newdata = pred_grid, type = "response")

ggplot(pred_grid, aes(x = HEIGHTM_KM, y = prob, color = factor(O2USED))) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("FALSE" = "tomato", "TRUE" = "steelblue"),
                     labels = c("No O2", "O2 Used")) +
  labs(title = "Predicted Summit Success Probability by Height and O2 Use",
       x = "Peak Height (km)",
       y = "Predicted Probability of Success",
       color = "Oxygen")
```
