---
title: "Supplemental Oxygen, Altitude, and Himalayan Expedition Success"
subtitle: "STAT 3302: Statistical Modeling for Discovery II"
author:
  - "Steven Dahms"
  - "Bardaan Dhesi"
  - "Kritarsh Negi"
group: "Himalayan B"
date: today
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(knitr)

options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

fmt_pct <- function(x, digits = 1) paste0(round(100 * x, digits), "%")
fmt_p <- function(x) {
  ifelse(x < 0.001, formatC(x, format = "e", digits = 2), sprintf("%.3f", x))
}

exped_tidy <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/exped_tidy.csv",
  show_col_types = FALSE
)

peaks_tidy <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-01-21/peaks_tidy.csv",
  show_col_types = FALSE
)

exped_model <- exped_tidy |>
  mutate(
    SUCCESS = as.integer(TERMREASON %in% 1:3),
    SEASON_FACTOR = factor(
      SEASON_FACTOR,
      levels = c("Autumn", "Spring", "Summer", "Winter")
    )
  ) |>
  select(
    PEAKID, YEAR, SUCCESS, TERMREASON, O2USED, SEASON_FACTOR,
    CAMPS, TOTMEMBERS
  ) |>
  left_join(
    peaks_tidy |> select(PEAKID, PKNAME, HEIGHTM, REGION_FACTOR),
    by = "PEAKID"
  ) |>
  mutate(
    HEIGHTM_KM = HEIGHTM / 1000,
    HEIGHTM_KM_C = HEIGHTM_KM - mean(HEIGHTM_KM, na.rm = TRUE)
  ) |>
  filter(
    !is.na(SUCCESS), !is.na(O2USED), !is.na(SEASON_FACTOR),
    !is.na(CAMPS), !is.na(TOTMEMBERS), !is.na(HEIGHTM_KM_C)
  )

model <- glm(
  SUCCESS ~ O2USED * HEIGHTM_KM_C + I(HEIGHTM_KM_C^2) +
    SEASON_FACTOR + CAMPS + TOTMEMBERS,
  data = exped_model,
  family = binomial
)

model_additive <- glm(
  SUCCESS ~ O2USED + HEIGHTM_KM_C + I(HEIGHTM_KM_C^2) +
    SEASON_FACTOR + CAMPS + TOTMEMBERS,
  data = exped_model,
  family = binomial
)

model_linear <- glm(
  SUCCESS ~ O2USED * HEIGHTM_KM_C + SEASON_FACTOR + CAMPS + TOTMEMBERS,
  data = exped_model,
  family = binomial
)

interaction_test <- anova(model_additive, model, test = "Chisq")
quadratic_test <- anova(model_linear, model, test = "Chisq")

n_expeditions <- nrow(exped_model)
n_peaks <- n_distinct(exped_model$PEAKID)
overall_success <- mean(exped_model$SUCCESS)
mean_height_km <- mean(exped_model$HEIGHTM_KM)
interaction_p <- interaction_test$`Pr(>Chi)`[2]
quadratic_p <- quadratic_test$`Pr(>Chi)`[2]
pearson_dispersion <- sum(residuals(model, type = "pearson")^2) /
  df.residual(model)
```

## Abstract

We studied whether supplemental oxygen and peak height predict the probability
that a Himalayan expedition successfully reaches a summit. Using 882 expeditions
from 2020-2024, we fit a logistic regression model with oxygen use, centered peak
height, a quadratic height term, season, number of high camps, team size, and an
oxygen-by-height interaction. The model shows that oxygen use is strongly
associated with summit success, and the oxygen advantage becomes much larger as
peak height increases.

## Introduction

Himalayan climbing is a high-risk activity where weather, altitude, preparation,
team size, and expedition strategy all affect the chance of success. Our main
research question is: what factors predict whether an expedition successfully
summits a Himalayan peak? We are especially interested in whether the effect of
supplemental oxygen depends on peak height.

This interaction is scientifically motivated. As altitude increases, lower
oxygen availability makes movement and decision-making more difficult. If
supplemental oxygen is useful, it should be most useful on the tallest peaks,
where hypoxia is most severe. At the same time, very high peaks may attract more
prepared and better-funded teams, so the raw relationship between height and
success may not be a simple downward trend.

## Data

The data come from the Himalayan Database through the 2025-01-21 TidyTuesday
release. The unit of observation is one expedition. The analysis uses
`r n_expeditions` expeditions from 2020-2024 across `r n_peaks` distinct peaks.
The response variable is `SUCCESS`, coded 1 when `TERMREASON` indicates success
on the main peak, subpeak, foresummit, or a claimed success, and 0 otherwise.

The overall success rate was `r fmt_pct(overall_success)`. Peak heights in the
model ranged from `r min(exped_model$HEIGHTM)` to `r max(exped_model$HEIGHTM)`
meters, with an average of `r round(mean_height_km, 2)` km. Oxygen use is not
randomly assigned: expeditions using oxygen were generally on taller peaks, so
we control for peak height directly in the model.

```{r}
#| label: tbl-oxygen-summary
exped_model |>
  group_by(O2USED) |>
  summarise(
    Expeditions = n(),
    `Success rate` = fmt_pct(mean(SUCCESS)),
    `Average height (km)` = round(mean(HEIGHTM_KM), 2),
    .groups = "drop"
  ) |>
  mutate(`Oxygen used` = if_else(O2USED, "Yes", "No"), .before = O2USED) |>
  select(-O2USED) |>
  kable(caption = "Expedition outcomes by supplemental oxygen use.")
```

```{r}
#| label: fig-height-success
#| fig-cap: "Initial height-success plot with individual expedition outcomes and a LOESS smooth."
height_success_plot <- exped_model |>
  mutate(
    outcome = factor(SUCCESS, levels = c(0, 1), labels = c("Failure", "Success"))
  )

ggplot(height_success_plot, aes(x = HEIGHTM_KM, y = SUCCESS)) +
  geom_jitter(
    aes(color = outcome),
    width = 0,
    height = 0.035,
    alpha = 0.25,
    size = 1.4
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    color = "black",
    fill = "gray70",
    linewidth = 1.1
  ) +
  scale_color_manual(values = c("Failure" = "tomato", "Success" = "steelblue")) +
  scale_y_continuous(
    breaks = c(0, 1),
    labels = c("Failure", "Success"),
    limits = c(-0.08, 1.08)
  ) +
  labs(
    x = "Peak height (km)",
    y = "Expedition outcome",
    color = "Outcome"
  )
```

This initial plot shows why a straight-line height effect is too simple. The
individual dots show each expedition at the failure and success margins, while
the smooth curve suggests a non-linear pattern: success drops at some high
altitudes but rises again among the tallest peaks. One plausible explanation is
self-selection, since the tallest peaks tend to be attempted by more experienced,
better-funded, and more prepared teams.However, due to the fact that most of the data was concentrated around roughy 5 peaks, we cannot rule out the possibility that the non-linear pattern is due to peak-level differences rather than a true height effect. We will return to this issue in the discussion.

## Methods

Because the response is binary, we model expedition success as a Bernoulli
outcome using logistic regression with a logit link. Let $\pi_i$ be the
probability that expedition $i$ succeeds. The fitted model is

$$
\log\left(\frac{\pi_i}{1-\pi_i}\right) =
\beta_0 + \beta_1 O2_i + \beta_2 H_i + \beta_3 H_i^2 +
\beta_4(O2_iH_i) + \beta_5 Season_i + \beta_6 Camps_i +
\beta_7 TeamSize_i,
$$

where $H_i$ is peak height in kilometers, centered at the sample mean. Height was
converted from meters to kilometers so a one-unit change is meaningful, and it
was centered to make the intercept and interaction more stable. We included a
quadratic height term because the initial height-success plot bends rather than
following a constant upward or downward slope. In context, that pattern is
reasonable: very tall peaks are more dangerous, but they also attract more
selective teams. A likelihood ratio test confirms that the quadratic term
improves fit compared with the same model without it
($p = `r fmt_p(quadratic_p)`$).

The main term of interest is `O2USED:HEIGHTM_KM_C`, which tests whether the
association between supplemental oxygen and success changes as peaks get taller.
We compare this model to an additive model without the interaction using a
likelihood ratio test.

## Results

```{r}
#| label: tbl-model-results
coef_mat <- coef(summary(model))

or_table <- tibble(
  term = rownames(coef_mat),
  estimate = coef_mat[, "Estimate"],
  se = coef_mat[, "Std. Error"],
  p_value = coef_mat[, "Pr(>|z|)"]
) |>
  mutate(
    Predictor = recode(
      term,
      "O2USEDTRUE" = "Oxygen used",
      "HEIGHTM_KM_C" = "Height, centered km",
      "I(HEIGHTM_KM_C^2)" = "Height squared",
      "SEASON_FACTORSpring" = "Spring vs Autumn",
      "SEASON_FACTORSummer" = "Summer vs Autumn",
      "SEASON_FACTORWinter" = "Winter vs Autumn",
      "CAMPS" = "High camps",
      "TOTMEMBERS" = "Team members",
      "O2USEDTRUE:HEIGHTM_KM_C" = "Oxygen x height"
    ),
    OR = exp(estimate),
    low = exp(estimate - 1.96 * se),
    high = exp(estimate + 1.96 * se),
    `95% Wald CI` = sprintf("%.2f to %.2f", low, high),
    `p-value` = fmt_p(p_value)
  ) |>
  filter(term != "(Intercept)") |>
  select(Predictor, OR, `95% Wald CI`, `p-value`)

or_table |>
  mutate(OR = round(OR, 2)) |>
  kable(caption = "Odds ratios from the fitted logistic regression model.")
```

At the centered height value, oxygen use is associated with much higher odds of
success (OR = 14.53, 95% CI 6.07 to 34.77). The centered height term is negative
(OR = 0.09), while the quadratic term is also significant, confirming that the
height effect is not constant over the range of the data.

The interaction between oxygen use and height is positive and significant
(OR = 27.85, $p = `r fmt_p(coef_mat["O2USEDTRUE:HEIGHTM_KM_C", "Pr(>|z|)"])`$).
The likelihood ratio test comparing the additive model to the interaction model
also supports keeping the interaction ($p = `r fmt_p(interaction_p)`$); AIC drops
from `r round(AIC(model_additive), 2)` to `r round(AIC(model), 2)`. This means the
estimated oxygen advantage increases sharply as peak height increases.

High camps and team size were also positively associated with success. Each
additional high camp multiplies the odds of success by about 1.77, and each
additional team member multiplies the odds by about 1.10. After controlling for
the other predictors, Spring expeditions have lower odds of success than Autumn
expeditions.

```{r}
#| label: fig-interaction
#| fig-cap: "Predicted summit success probability by peak height and oxygen use, holding season at Autumn and numeric controls at their sample means."
height_mean <- mean(exped_model$HEIGHTM_KM, na.rm = TRUE)

pred_grid <- expand.grid(
  HEIGHTM_KM = seq(
    min(exped_model$HEIGHTM_KM),
    max(exped_model$HEIGHTM_KM),
    length.out = 150
  ),
  O2USED = c(FALSE, TRUE),
  SEASON_FACTOR = factor("Autumn", levels = levels(exped_model$SEASON_FACTOR)),
  CAMPS = mean(exped_model$CAMPS, na.rm = TRUE),
  TOTMEMBERS = mean(exped_model$TOTMEMBERS, na.rm = TRUE)
) |>
  mutate(HEIGHTM_KM_C = HEIGHTM_KM - height_mean)

pred_grid$probability <- predict(model, newdata = pred_grid, type = "response")

ggplot(
  pred_grid,
  aes(x = HEIGHTM_KM, y = probability, color = O2USED)
) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(
    values = c("FALSE" = "tomato", "TRUE" = "steelblue"),
    labels = c("No oxygen", "Oxygen used")
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(
    x = "Peak height (km)",
    y = "Predicted probability of success",
    color = "Supplemental oxygen"
  )
```

## Diagnostics

We checked residual plots rather than relying only on a calibration plot.
For logistic regression, residuals do not need to be normally distributed, but
they should be centered around zero without a strong trend across fitted
probabilities. The Pearson dispersion estimate was
`r round(pearson_dispersion, 2)`, which does not suggest serious overdispersion.

```{r}
#| label: fig-deviance-residuals
#| fig-cap: "Deviance residuals plotted against fitted summit-success probabilities."
resid_diag <- exped_model |>
  mutate(
    fitted_prob = fitted(model),
    deviance_resid = residuals(model, type = "deviance"),
    pearson_resid = residuals(model, type = "pearson")
  )

ggplot(resid_diag, aes(x = fitted_prob, y = deviance_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, color = "tomato", linewidth = 1) +
  labs(
    x = "Fitted probability of summit success",
    y = "Deviance residual"
  )
```

```{r}
#| label: fig-binned-pearson-residuals
#| fig-cap: "Average Pearson residual by fitted-probability decile."
resid_diag |>
  mutate(prob_decile = ntile(fitted_prob, 10)) |>
  group_by(prob_decile) |>
  summarise(
    mean_fitted = mean(fitted_prob),
    mean_pearson_resid = mean(pearson_resid),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = mean_fitted, y = mean_pearson_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_line(color = "steelblue") +
  geom_point(aes(size = n), color = "steelblue") +
  guides(size = "none") +
  labs(
    x = "Average fitted probability by decile",
    y = "Average Pearson residual"
  )
```

The residual plots are mostly centered around zero, suggesting that the model
captures the main success-probability pattern. There is still some structure in
the middle fitted-probability range, which is consistent with missing predictors
such as route difficulty, short-term weather, expedition support, and climber
experience.

## Discussion

The results support the central hypothesis: supplemental oxygen is not only
associated with higher summit success overall, but its association with success
is stronger on taller peaks. This matches the physiological motivation that
oxygen matters more when altitude-related hypoxia is more severe.

The findings should not be interpreted as a randomized causal effect. Teams
choose whether to use oxygen, and that choice is related to peak height,
resources, experience, route difficulty, and expedition goals. We controlled for
height, season, camps, and team size, but the dataset does not fully capture
climber skill, weather severity, budget, Sherpa support, or route-level technical
difficulty. There may also be repeated dependence among expeditions on the same
popular peaks, especially Mount Everest and other 8000-meter peaks.

With more time, we would extend the analysis with peak-level random effects or
fixed effects for heavily climbed peaks, add route and weather information, and
separate main-peak success from subpeak or claimed success. We would also check
whether oxygen use is acting partly as a proxy for expedition resources.

## Conclusion

For Himalayan expeditions from 2020-2024, summit success is strongly associated
with supplemental oxygen, peak height, season, number of high camps, and team
size. The most important result is the positive oxygen-by-height interaction:
oxygen use appears increasingly beneficial as peak height rises. This suggests
that planning for extreme-altitude climbs should treat oxygen strategy and peak
height as connected factors rather than separate predictors.

## References

The Himalayan Database. https://www.himalayandatabase.com/

R for Data Science TidyTuesday. 2025-01-21 Himalayan Expeditions data.
https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-01-21

## Appendix {.appendix}

```{r}
#| label: appendix-model-comparisons
anova(model_additive, model, test = "Chisq")
AIC(model_additive, model)

anova(model_linear, model, test = "Chisq")
AIC(model_linear, model)
```
